Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-19 paper/main.tex
update section 4

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.jpeg}
    \caption{95\% credible intervals for the power spectrum of the first Mira-Titan
    cosmology (centered by the predicted posterior mean). Perturbation theory, 
    low-resolution, and high-resolution runs are shown for context.}
    \label{fig:plot_fit}
\end{figure}

@@

Here, we begin with a $n \times m$ matrix composed of $m=111$ posterior means, 
each of length $n$, denoted by $\boldsymbol\eta$. To facilitate more efficient estimation, 
we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, 
where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ 
and $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. 
For this application, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}=111$, which 
determines the number of non-zero elements in $D$. We find an alternative decomposition 
where $\Sigma$ is a diagonal matrix with the non-zero singular values of $\boldsymbol\eta$, 
and $U_1$ and $V_1$ are matrices with orthonormal columns of size $n_\eta \times r$ 
and $m \times r$, respectively.

Following \cite{higdon2008computer, higdon2010estcosmo}, we equivalently decompose 
$\boldsymbol\eta$ into a principal component (PC) basis matrix 
$B^* = \frac{1}{\sqrt{m}}U_1\Sigma$ along with its corresponding weights 
$W^* = \sqrt{m}V_1$. Without loss of generality, we can perform this decomposition 
after an overall mean trend is removed. If we use $p_\eta < r$ PCs, this will reduce 
$B$ and $W$ to be the first $p_\eta$ columns of $B^*$ and $W^*$ respectively and 
result in an approximation where

\begin{equation}
    \boldsymbol\eta= UDV^T = U_1\Sigma V_1^T \approx BW^T.
\end{equation}

Our goal is to predict the power spectrum for a held-out cosmology. 
For simplicity, we will denote these eight parameters for a given cosmology (see 
Section \ref{sec:data}) as $\psi \in \mathbb{R}^8$. 
Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated 
mean trend, the principal components contained in $B$, and one example of how the 
weight for the first PC will vary dependent on the first of $p_\psi=8$ cosmological 
parameters contained in $\psi$.

In the following subsections, we consider different modeling strategies for $W$, 
which will influence predictions and sensitivity analyses based on $\boldsymbol\eta$.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{The estimated mean trend of the 111 posterior means (left), the principal 
    components contained in $B$ (middle), and an illustration of how the weights 
    for the first PC will change as the first cosmological parameter varies (right) 
    \textcolor{red}{This last plot is wonky because it was made with only the 6 test 
    cosmologies, not a large uniform or FF design, Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

@@

Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

@@

From our 32 posterior means using DGP.FCO, these are input into SEPIA in order to predict power spectra
for the 32 test cosmologies. With the infinite-resolution output, we have a sense of the ``best-case" 
predictions we can have (i.e., if we trained with the 32 true power spectra, as opposed to our 
estimated posterior means). In Figure \ref{fig:pca_preds_v_camb}, we see comparisons of the heldout 
infinite resolution alongside two predictions: one based on DGP.FCO and the ``best-case" scenario using the
infinite-resolution spectra of the training set. There is a high level of agreement between the pairs
of predictions across cosmologies. Upon comparing MSEs for each of these prediction scenarios 
(Figure \ref{fig:mse_camb}, most are very similar across the two groups (with the two most difficult 
predictions being the exceptions), suggesting that overall DGP.FCO is estimating the true spectra 
for each batch well.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions for 16 of the 32 test cosmologies after centering and scaling. 
             The interpolated infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue and the prediction from 
             the DGP.FCO posterior mean is in red.}   
    \label{fig:pca_preds_v_camb}
\end{figure}

@@

Here we show prediction results for 6 held-out cosmologies. We compare our method 
with Cosmic Emu \citep{moran2023mira}, the emulator constructed using the full suite 
of 111 Mira-Titan model runs. Cosmic Emu uses a Bayesian approach where the smooth 
spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep{moran2024dpc}.

Both Cosmic Emu and DGP.FCO are capable of modeling a true underlying functions based 
upon multiple functional realizations, but our approach utilizes DGPs within a 
hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
smooth spectra from the training cosmologies, each method utilizes GPs in order to 
estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
performance between Cosmic Emu and DGP.FCO for the 6 held-out cosmologies. 
\textcolor{magenta}{In addition, we can also predict over a large uniform or full 
factorial design to obtain estimated main effects for each cosmological parameter in 
$\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
variation by main effects, two-way interactions, and so on. Should this be included?}

Plots for the average root mean squared error (RMSE) for each prediction method over the 
6 cosmologies and each $k$ value are shown in Figure \ref{fig:plot_rmse_k}. Comparisons 
of the spectrum estimation for training cosmologies are shown as well (using RMSE against 
the weighted average of observations). For predictions, DGP.FCO achieves the greatest 
reduction in RMSE in the region where perturbation theory and the low-resolution runs 
are deemed unbiased. For the region where only the high-resolution run is unbiased, 
Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.FCO has 
an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is 
lower at 56\% of $k$ values considered. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{rmse_by_k.png}
    \caption{Left: Average RMSE of each prediction method across all $k$ values. 
             Right: Average RMSE for each method when using the held-out computer 
             model runs to train the model. In each case, the truth is assumed to 
             be the weighted average of the computer model runs.}
    \label{fig:plot_rmse_k}
\end{figure}

@@

\textcolor{magenta}{The different appendices from dissertation as necessary. 
Can also put details of dataset in supplementary material.} 
\textcolor{red}{Fix references as necessary, e.g. Moran et al. 2024.}

@@

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{plot_fit.jpeg}
    \caption{{\it Left:} 95\% credible intervals for the power spectrum of the first Mira-Titan
    cosmology (centered by the predicted posterior mean). Perturbation theory, 
    low-resolution, and high-resolution runs are shown for context. {\it Right:} Posterior draws
    of $W$, illustrating some nonstationary behavior.}
    \label{fig:plot_fit}
\end{figure}

@@

Our goal is to predict the power spectrum for a held-out cosmology. 
For simplicity, we will denote these eight parameters for a given cosmology (see 
Section \ref{sec:data}) as $\psi \in \mathbb{R}^{p_\psi}$. With our estimated posterior means
from DGP.FCO for each training cosmology, we now predict $S$ for unobserved cosmologies
as a function of $\psi$.

We have $m$-many training observations ($j = 1,\dots,m$) which each contain the $p_\psi$-dimensional
input $\psi_j$ and DGP.FCO estimates $\hat{S}_j$ for the $j$th cosmology. 
We combine all $\hat{S}_j$ estimates into a $n\times m$ matrix $\boldsymbol\eta$.
Due to the fact that each column of $\boldsymbol\eta$ represents functional output, 
we choose to reformulate the problem using principal components (PCs). Details of this can be found in 
\ref{sec:apdx_basis}. Here we will use the first $p_\eta$ PCs.

Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated 
mean trend for the Mira-Titan dataset, the $p_\eta$ basis functions, and one 
example of how the weight for the first PC will vary dependent on the first of 
$p_\psi=8$ cosmological parameters contained in $\psi$. As a result, our goal becomes 
estimating the weights $\Gamma$ corresponding to each of the $p_\eta$ PCs, 
as a function of $\psi$.

@@

Using these trained GPs from the PCs, we predict the spectra for each of the $N$ held-out 
cosmologies for both the CAMB and Mira-Titan datasets.

@@

From our $m=32$ posterior means using DGP.FCO, these are input into SEPIA in order to predict power spectra
for the $N=32$ test cosmologies. In the CAMB dataset, $p_\psi=6$. With the infinite-resolution spectra, we 
have a sense of the ``best-case" predictions we can have (i.e., if we trained with the 32 true 
infinite-resolution spectra, as opposed to our estimated posterior means). In Figure \ref{fig:pca_preds_v_camb}, 
we see comparisons of the heldout infinite resolution alongside two predictions: one 
based on DGP.FCO and the ``best-case" scenario using the infinite-resolution spectra of 
the training set. There is a high level of agreement between the pairs of predictions 
across cosmologies. Upon comparing MSEs for each of these prediction scenarios 
(Figure \ref{fig:mse_camb}, most are very similar across the two groups (with the two most difficult 
predictions being the exceptions), suggesting that overall DGP.FCO is estimating the true spectra 
for each batch well.

\begin{figure}
    \centering
    \includegraphics[width=4in]{preds_diff1_4_8.jpg}
    \caption{Predictions for 16 of the $N=32$ test cosmologies after centering and scaling. 
             The interpolated infinite resolution run is shown in black, the prediction 
             (trained on the infinite resolution) is in blue and the prediction from 
             the DGP.FCO posterior mean is in red.}   
    \label{fig:pca_preds_v_camb}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{{\it Left:} The estimated mean trend of the $m=111$ posterior means from 
    the Mira-Titan dataset. {\it Middle:} The $p_\eta$ principal components obtained from 
    the posterior means. {\it Right:} An 
    illustration of how the weights for the first PC will change as the first 
    cosmological parameter varies \textcolor{red}{This last plot is wonky because 
    it was made with only the $N=6$ test cosmologies, not a large uniform or FF design, 
    Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

Here we show prediction results for $N=6$ held-out cosmologies for the Mira-Titan dataset. 
We compare our method with Cosmic Emu \citep{moran2023mira}, the emulator constructed using the full suite 
of $m=111$ Mira-Titan model runs. Cosmic Emu uses a Bayesian approach where the smooth 
spectrum for each cosmology is modeled with a process convolution on Brownian motion. 
Nonstationarity is permitted through modeling the bandwidth parameter with a process 
convolution as well; this composition is referred to as a deep process convolution 
\citep{moran2024dpc}.

Both Cosmic Emu and DGP.FCO are capable of modeling a true underlying functions based 
upon multiple functional realizations, but our approach utilizes DGPs within a 
hierarchical model in lieu of deep process convolutions. After obtaining the estimated 
smooth spectra from the training cosmologies, each method utilizes GPs in order to 
estimate the weights of the PCs from the estimated spectra. Below, we compare prediction 
performance between Cosmic Emu and DGP.FCO for the $N=6$ held-out cosmologies. 
% \textcolor{magenta}{In addition, we can also predict over a large uniform or full 
% factorial design to obtain estimated main effects for each cosmological parameter in 
% $\psi$, interactions amongst the $p_\psi=8$ parameters, and decompose the overall 
% variation by main effects, two-way interactions, and so on. Should this be included?}

Plots for the average root mean squared error (RMSE) for each prediction method over the 
$N=6$ cosmologies and each $k$ value are shown in Figure \ref{fig:plot_rmse_k}. Comparisons 
of the spectrum estimation for training cosmologies are shown as well (using RMSE against 
the weighted average of observations). For predictions, DGP.FCO achieves the greatest 
reduction in RMSE in the region where perturbation theory and the low-resolution runs 
are deemed unbiased. For the region where only the high-resolution run is unbiased, 
Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.FCO has 
an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is 
lower at 56\% of $k$ values considered. The individual predictions for each method and 
cosmology are shown in Figure \ref{fig:plot_pred_1to6}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=6in]{rmse_by_k.png}
    \caption{{\it Left:} Average RMSE of each prediction method across all $k$ values. 
             {\it Right:} Average RMSE for each method when using the held-out computer 
             model runs to train the model. In each case, the truth is assumed to 
             be the weighted average of the computer model runs.}
    \label{fig:plot_rmse_k}
\end{figure}

@@

% \textcolor{magenta}{Steve The different appendices from dissertation as necessary. 
% Can also put details of dataset in supplementary material.} 
% \textcolor{red}{Fix references as necessary, e.g. Moran et al. 2024.}

@@

\section{Details on Basis Decomposition for Predictions}
\label{sec:apdx_basis}

Here, we begin with a $n \times m$ matrix composed of $m=111$ posterior means, 
each of length $n$, denoted by $\boldsymbol\eta$. To facilitate more efficient estimation, 
we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, 
where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ 
and $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. 
For this application, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}$ (111 for Mira-Titan, 
32 for CAMB), which determines the number of non-zero elements in $D$. We find an alternative 
decomposition where $\Sigma$ is a diagonal matrix with the non-zero singular values of 
$\boldsymbol\eta$, and $U_1$ and $V_1$ are matrices with orthonormal columns of size 
$n_\eta \times r$ and $m \times r$, respectively.

Following \cite{higdon2008computer, higdon2010estcosmo}, we equivalently decompose 
$\boldsymbol\eta$ into a principal component (PC) basis matrix 
$B^* = \frac{1}{\sqrt{m}}U_1\Sigma$ along with its corresponding weights 
$\Gamma^* = \sqrt{m}V_1$. Without loss of generality, we can perform this decomposition 
after an overall mean trend is removed. If we use $p_\eta < r$ PCs, this will reduce 
$B$ and $\Gamma$ to be the first $p_\eta$ columns of $B^*$ and $\Gamma^*$ respectively and 
result in an approximation where

\begin{equation}
    \boldsymbol\eta= UDV^T = U_1\Sigma V_1^T \approx B\Gamma^T.
\end{equation}
